Informing policy via dynamic models: Cholera in Haiti

### 2024-02-29 ms.Rnw
Minor edits for clarity

if (RUN_LEVEL == 3) {
  # Print session info into a txt file. Only update if RUN_LEVEL == 3, which
  # is when the final results are calculated.
  cat(
    utils::capture.output(sessionInfo()),
    file = "sessionInfo_tmp.txt",
    sep="\n"
  )
}

@@

\section*{Author summary}
Quantitative understanding of infectious disease transmission dynamics relies upon mathematical models informed by scientific knowledge and relevant data.
The models aim to provide a statistical description of the trajectory of an epidemic and its uncertainty, together with a representation of the underlying biological mechanisms.
Evaluation of success at these goals is necessary in order for a model to provide a reliable tool for guiding evidence-based public policy interventions.
\old{In this article, we discuss common modeling decisions that may not provide an adequate statistical explanation of the data, and we show that attention to statistical fit can lead to more accurate policy evaluations}.
\new{In this article, we conduct a re-analysis of the 2010 cholera outbreak in Haiti. We use this case study to investigate current procedures for fitting mathematical models to time series data, while identifying limitations of these methodologies and proposing remedies.}
\new{We present} \old{Our analysis presents} methodology for diagnosing how well a model describes observed data\new{.}\old{, and provides careful justification of valid conclusions from the fitted model.}
Objective measures are used to \old{benchmark} \new{evaluate} model fit; when \old{these}\new{objective measures of the goodness-of-fit of a model} are combined with reproducibility, a framework emerges for continual improvement \new{to model based inference} when revisiting the data\old{ and models}.
\new{This framework promotes scientific discovery and supports model-based policy recommendations by simplifying the process for researchers to build off of previous results.}
Our data analysis workflow is supported by recent advances in algorithms, software and hardware, which facilitate statistical fitting of nonlinear stochastic dynamic models to observed incidence data.
However, inference for high-dimensional systems remains a methodological challenge.
One of the models under consideration involves spatially coupled stochastic meta-populations, and we demonstrate how a recently developed algorithm permits likelihood-based inference and model diagnostics in this setting.
We contend that raising the currently accepted standards of infectious disease modeling will result in a greater ability of scientists and policy makers to understand and respond to future infectious disease outbreaks.

@@

This case study investigates the use of dynamic models and spatiotemporal data to inform \old{a policy decision} \new{public health policy} in the context of the cholera outbreak in Haiti, which started in 2010.
\old{We build on a multi-group modeling exercise by Lee et al. [1] in which four expert modeling teams developed models to the same dataset with the goal of comparing conclusions on the feasibility of eliminating cholera by a vaccination campaign}.
\old{The data that were used for this study, and that is the subject of our reanalysis, is displayed in Fig.~[1].}
\new{Various dynamic models were developed to study this outbreak: searching PubMed with keywords ``Haiti, cholera, model'' we obtained 22 studies that utilized deterministic mechanistic dynamic models \cite{lee20,tuite11,andrews11,botelho21,fitzgibbon20,eisenberg13,rinaldo12,chao11,date11,lin19,abrams13,collins21,akman15,trevisin22,mavian20,collins14,kelly16,capone15,leung22,mari15,gatto12,kuhn14}
and 11 studies that used stochastic models \cite{kirpich16,lee20,pasetto18,mukandavire13,kirpich17,lewnard16,kunkel17,mukandavire15,sallah17,azman12,azman15}.
Incidence data on the outbreak are available at various spatial scales, motivating 17 studies in our literature review to consider spatially explicit dynamic models \cite{lee20,tuite11,pasetto18,fitzgibbon20,eisenberg13,rinaldo12,chao11,abrams13,trevisin22,sallah17,collins14,kelly16,azman12,leung22,kuhn14,mari15,gatto12}.
Here we focus on a multi-group modeling exercise by Lee et al.~\cite{lee20} in which four expert modeling teams developed models to the same dataset with the goal of comparing conclusions on the feasibility of eliminating cholera by a vaccination campaign.}
Model~1 is stochastic and describes cholera at the national level;
Model~2 is deterministic with spatial structure, and includes transmission via contaminated water;
Model~3 is stochastic with spatial structure, and accounts for measured rainfall.
Model~4 has an agent-based construction, featuring considerable mechanistic detail but limited ability to calibrate these details to data.
\new{These modeling strategies were selected by \cite{lee20} to represent the range of approaches used in the research community}
We focus on Models~1--3, as the strengths and weaknesses of the agent-based modeling approach \cite{tracy18} are outside the scope of this article.
\new{In addition, agent-based models were less widely used, the agent based model of \cite{lee20} being the only model of this class that was found in our literature review.}
\new{The data that were used in \cite{lee20}, and that we reanalyze, are displayed in Fig.~\ref{fig:CholeraData}.}

@@

The four independent teams were given the task of estimating the
potential effect of prospective oral cholera vaccine (OCV) programs.
While OCV is accepted as a safe and effective tool for controlling the spread of cholera, the global stockpile of OCV doses remains limited \cite{pezzoli20}.
Advances in OCV technology and vaccine availability, however, raised the possibility of planning a national vaccination program \cite{lee20}.
\new{The possibility of controlling the Haiti cholera outbreak via OCV was considered by various research groups \cite{ivers17,andrews11,walton11,matias17,chao11,fung13,date11,leung22,azman15}}.
In the \new{Lee et al.~\cite{lee20}} study, certain data were shared between the groups, including demography and vaccination history; vaccine efficacy was also fixed at a shared value between groups.
Beyond this, the groups made autonomous decisions on what to include and exclude from their models.
\old{The groups largely adhered to existing guidelines for creating models to inform policy [35, 36] and, d}\new{D}espite their autonomy, \new{the four independent teams} obtained a consensus that an extensive nationwide vaccination campaign would be necessary to eliminate cholera from Haiti.
\new{They also provided a range for the forecasted number of cumulative cases from February 2019 to Februaruy 2024 based on model simulations, with median estimates from each model measured in hundreds of thousands.}
\old{Their forecasts, however, were} \new{These forecasts are} inconsistent with the prolonged period with no confirmed cholera cases between February, 2019 and September, 2022 \cite{trevisin22}.
Though cholera has recently reemerged in Haiti \cite{rubin22,PAHO23}, the inability to accurately forecast cholera incidence from 2019-2022\old{---while diligently adhering to existing scientific standards for infectious disease modeling---demonstrates that existing standards of disease modeling and forecasting may be insufficient.} \new{prompts us to consider retrospectively what may have been done differently in order to obtain more reliable conclusions.}

@@

\new{Confidence intervals for model parameters are given in supplement~S8.}

@@

\subsubsection*{Model~1}
\label{sec:model1}
The latent state vector $\bm{X}(t) = \big(S_{\vaccCounter}(t),  E_{\vaccCounter}(t), I_{\vaccCounter}(t), A_{\vaccCounter}(t), R_{\vaccCounter}(t), \vaccCounter \in 0:\vaccClass\big)$ describes susceptible, latent (exposed), infected (and symptomatic), asymptomatic, and recovered individuals in vaccine cohort $\vaccCounter$ at time $t$.
Here, $\vaccCounter=0$ corresponds to unvaccinated individuals, and $\vaccCounter \in \seq{1}{\vaccClass}$ describes hypothetical vaccination programs\new{, where each program $\vaccCounter$ indexes differences in both the number of doses administered (one versus two doses per individual) and the round of vaccine administration.}
The force of infection is
\begin{equation}
\label{model1:lambda}
\lambda(t) = \Big(\sum_{\vaccCounter=0}^{\vaccClass}  I_{\vaccCounter}(t) + \asymptomRelativeInfect \sum_{\vaccCounter=0}^{\vaccClass} A_{\vaccCounter}(t) \Big)^\nu \frac{d\Gamma(t)}{dt} \transmission(t)/N,
\end{equation}
where $\transmission(t)$ is a periodic cubic spline representation of seasonality, given in terms of a B-spline basis $\{ s_j(t), j\in \seq{1}{6}\}$ and parameters $\transmission_{1:6}$ as
\begin{equation}
\label{model1:beta}
\log\transmission(t) = \sum_{j=1}^6 \transmission_j s_j(t).
\end{equation}
The process noise $d\Gamma(t)/dt$ is multiplicative Gamma-distributed white noise, with infinitesimal variance parameter $\sigmaProc^2$.
Lee et al.~\cite{lee20} included process noise in Model~3 but not in Model~1, i.e., they fixed $\sigmaProc^2 = 0$.
Gamma white noise in the transmission rate gives rise to an over-dispersed latent Markov process \cite{breto11} which has been found to improve the statistical fit of disease transmission models \cite{stocks20,he10}.

@@

Individuals move from department $u$ to $v$ at rate $T_{uv}$, and aquatic cholera moves at rate $T^W_{uv}$.
The nonzero transition rates are
\begin{eqnarray}
\label{model2:mu_SE}
\mu_{S_{u\vaccCounter}E_{u\vaccCounter}} &=& \new{(1 - \vaccineEfficacy_\vaccCounter) \, \lambda_u(t)},
\\
\label{model2:mu_EI}
\mu_{E_{u\vaccCounter}I_{u\vaccCounter}} &=& \symptomFrac\muEI, \quad \mu_{E_{u\vaccCounter}A_{u\vaccCounter}} = (1-\symptomFrac)\muEI,
\\
\label{model2:mu_IR}
\mu_{I_{u\vaccCounter}R_{u\vaccCounter}} &=& \mu_{A_{u\vaccCounter}R^A_{u\vaccCounter}} = \muIR,
\\
\label{model2:RS}
\mu_{R_{u\vaccCounter}S_{u\vaccCounter}} &=& \mu_{R^A_{u\vaccCounter}S_{u\vaccCounter}} = \muRS,
\\
\label{model2:transport}
\mu_{S_{u\vaccCounter}S_{{\varv}\vaccCounter}} &=& \mu_{E_{u\vaccCounter}E_{{\varv}\vaccCounter}} = \mu_{I_{u\vaccCounter} I_{{\varv}\vaccCounter}} = \mu_{A_{u\vaccCounter}A_{{\varv}\vaccCounter}} = \mu_{R_{u\vaccCounter}R_{{\varv}\vaccCounter}} = \mu_{R^A_{u\vaccCounter} R^A_{{\varv}\vaccCounter}} = T_{u\varv},
\\
\label{model2:omega1}
\mu_{S_{u1}S_{u0}} &=& \mu_{S_{u3}S_{u0}} = \omega_1,
\\
\label{model2:omega2}
\mu_{S_{u2}S_{u0}} &=& \mu_{S_{u4}S_{u0}} = \omega_2,
\\
\label{model2:to_W}
\mu_{\demography W_u} &=& \Wshed \left\{ \sum_{\vaccCounter=0}^4 I_{u\vaccCounter}(t) + \asymptomRelativeShed \sum_{\vaccCounter=0}^4 A_{u\vaccCounter}(t) \right\},
\\
\label{model2:from_W}
\mu_{W_u\demography} &=& \Wremoval,
\\
\label{model2:water_transport}
\mu_{W_uW_{\varv}} &=& w_r T^W_{u\varv}.
\end{eqnarray}
In Eq.~\myeqref{model2:transport} the spatial coupling is specified by a gravity model,
\begin{equation}
\label{model2:gravity}
T_{u\varv} = v_{\mathrm{rate}} \times \frac{\mathrm{Pop}_u \mathrm{Pop}_{\varv}}{D_{u\varv}^2},
\end{equation}
where $\mathrm{Pop}_u$ is the mean population for department $u$,
$D_{u\varv}$ is a distance measure estimating average road distance between randomly chosen members of each population, and $v_{\mathrm{rate}}= 10^{-12} \,\mbox{km$^2$yr$^{-1}$}$ \old{was treated as a fixed constant.}\new{was fixed at the value used by \cite{lee20}.}
In Eq.~\myeqref{model2:water_transport}, $T^W_{u\varv}$ is a measure of river flow between departments.
The unit of $W_u(t)$ is cells per ml, with dose response modeled via a saturation constant of $\Wsat$ in Eq.~\myeqref{model2:lambda}.
In Eq.~\myeqref{model2:mu_SE}, $\vaccineEfficacy_\vaccCounter$ denotes the vaccine efficacy for each vaccination campaign $\vaccCounter \in \vaccClass$, with \old{$\vaccineEfficacy_0 = 1$, $\vaccineEfficacy_1 = 1 - 0.429 \childReduce$, $\vaccineEfficacy_2 = 1 - 0.519 \childReduce$, $\vaccineEfficacy_3 = 1 - 0.429$, and $\vaccineEfficacy_4 = 1 - 0.519$} \new{$\vaccineEfficacy_0 = 0$, $\vaccineEfficacy_1 = 0.429 \childReduce$, $\vaccineEfficacy_2 = 0.519 \childReduce$, $\vaccineEfficacy_3 = 0.429$, and $\vaccineEfficacy_4 = 0.519$}
Here, $\childReduce = 0.4688$ represents the reduced efficacy of the vaccination for children under the age of five years, and the values $0.429$ and $0.519$ are the median effectiveness of one and two doses over their effective period respectively, according to Table S4 in the supplement material of Lee et al.~\cite{lee20}.
Because vaccine efficacy remains constant, individuals in this model transition from a vaccinated compartment to the susceptible compartment at the end of the vaccine coverage period.

@@

\begin{table}[!h]
\centering
\caption{\label{tab:likes}{\bf AIC values for Models~1--3 and their benchmarks.}}
\begin{tabular}{|c|c|c|c|}
\hline
 & \thead{Model~1} & \thead{Model~2} & \thead{Model~3}
\\
\hline
\hline
\multirow{2}{*}{Log-likelihood} &
  $\Sexpr{myround(mod1_ll, 1)}$ &
  $\Sexpr{myround(mod2_ll, 1)}$ &
  $\Sexpr{myround(mod3_ll, 1)}$ \\
    & ($\Sexpr{myround(mod1_lee_ll, 1)}$)\footnotemark[1] &
  ($\Sexpr{myround(unname(lee20_replication['leeFit_ll']), 1)}$) &
  ($\Sexpr{myround(lee3_results[["all_ll_eval"]], 1)}$)\footnotemark[2]
\\
\hline
Number of &
  $\Sexpr{as.character(mod1_n_params)}$ &
  $\Sexpr{as.character(mod2_n_params)}$ &
  $\Sexpr{as.character(mod3_n_params)}$ \\
 Fit Parameters &
 (\Sexpr{as.character(mod1_lee_n_params)}) &
 ($\Sexpr{as.character(unname(lee20_replication['leeFit_n_params']))}$) &
 ($\Sexpr{as.character(lee3_results$n_params, 1)}$)
\\
\hline
\multirow{2}{*}{AIC} &
  $\Sexpr{myround(mod1_aic, 1)}$ &
  $\Sexpr{myround(mod2_aic, 1)}$ &
  $\Sexpr{myround(mod3_aic, 1)}$ \\
  & ($\Sexpr{myround(mod1_lee_aic, 1)}$)\footnotemark[1] &
  ($\Sexpr{myround(mod2_lee_aic, 1)}$) &
  ($\Sexpr{myround(mod3_lee_coupled_aic, 1)}$)\footnotemark[2]
\\
\hline
\thead{Benchmark \\ AIC} &
  $\Sexpr{myround(-2 * benchmarks$m1 + 2 * 4, 1)}$ &
  $\Sexpr{myround(-2 * benchmarks$m2 + 2 * 40, 1)}$ &
  $\Sexpr{myround(-2 * benchmarks$m3 + 2 * 40, 1)}$
\\
\hline
\end{tabular}
\begin{flushleft}
Values in parentheses are corresponding values using Lee et al.~\cite{lee20} parameter estimates.  \textsuperscript{1}The reported likelihood is an upper bound of the likelihood of the Lee et al.~\cite{lee20} model. \textsuperscript{2}Lee et al.~\cite{lee20} fit Model~3 to a subset of the data (March 2014 onward, excluding data from Ouest in 2015-2016).
On this subset, their model has a likelihood of $\Sexpr{myround(lee3_results$sub_ll_eval, 1)}$.
On this same subset, our model has a likelihood of $\Sexpr{myround(mod3_eval_res$subset_ll['est'], 1)}$.
Details of estimating the likelihood of the Lee et al.~\cite{lee20} models is provided in \nameref{S_lee20}
See Sec.~S6 of the supplement material for more details on estimating the likelihood of the Lee et al.~\cite{lee20} models.
\end{flushleft}
\end{table}

@@

% One of the most frequently used approaches to calibrate the parameters of a statistical model is maximum likelihood estimation.
% In particular, a closed form expression of the joint model density---described in Eq.~\myeqref{eq:jointLik}---is not readily available, making likelihood based inference a challenging task.
% In the case of any POMP model, the likelihood function---as described in Eq.~\myeqref{eq:likedef}---is written as $\mathcal{L}(\paramVec) = \int f_{\bm{X}_{0:N}, \bm{Y}_{1:N}}\big(\bm{x}_{0:N}, \bm{y}_{1:N};\paramVec\big) d\bm{x}_{0:N}$.
% The high-dimensional integral in the likelihood, and the potentially non-linear nature of the process model, make it impossible to obtain a closed form expression of the likelihood function of POMP models except in the most simple cases.
% Model~1 was developed with the intent of providing a description of cholera dynamics that is scientifically plausible rather than on that is simply statistically convenient.
% Because of this, a closed form expression of the likelihood of Model~1 is not readily available.

% In order to retain the ability to propose models that are scientifically meaningful rather than only those that are simply statistically convenient, we restrict ourselves to parameter estimation techniques that have the plug-and-play property, which is that the fitting procedure only requires the ability to simulate the latent process instead of evaluating transition densities \cite{breto09,he10};
% in the context of the notation and definitions employed in this article, this means that we only require the ability to simulate from $f_{\bm{X}_0}\big(\bm{x}_0;\paramVec\big)$ and $f_{\bm{X}_n|\bm{X}_{n-1}}\big(\bm{x}_{n}|\bm{x}_{n-1}; \paramVec\big)$ rather than needing to evaluate these densities.
% Plug-and-play algorithms include Bayesian approaches like ABC and PMCMC \cite{toni09,andrieu10}, but here we use algorithms that enable maximum likelihood estimation.
% To our knowledge, the only plug-and-play methods that can be used to maximize the likelihood for POMP models of this complexity are iterated filtering algorithms \cite{ionides15}, which modify the well-known \emph{particle filter} \cite{arulampalam02}.
% The particle filter is often used to obtain values of latent states, but can also be used to obtain likelihood estimates of a POMP model.
% Iterated filtering algorithms such as IF2 \cite{ionides15} extend the particle filter by performing a random walk for each parameter and particle;
% these perturbations are carried out iteratively over multiple filtering operations, using the collection of parameters from the previous filtering pass as the parameter initialization for the next iteration, and decreasing the random walk variance at each step.

@@

The ongoing global COVID-19 pandemic has \old{provided a clear example on} \new{demonstrated} how government policy may be affected by the \old{conclusions of scientific models} \new{inferences drawn from mathematical modeling} \cite{saltelli20}.
\old{This article demonstrates that fitting appropriate scientific models to guide policy is a challenging statistical task.}
\new{However, the development of credible models---which are supported by data and can lend quantitative insights into a dynamic system---remains a challenging task.
In this article, we demonstrate opportunites available for raising the current standards of statistical inference for mathematical models of biological systems.

>>>>

if (RUN_LEVEL == 3) {
  # Print session info into a txt file. Only update if RUN_LEVEL == 3, which
  # is when the final results are calculated.
  cat(
    utils::capture.output(sessionInfo()),
    file = "sessionInfo.txt",
    sep="\n"
  )
}

@@

\section*{Author summary}
Quantitative understanding of infectious disease transmission dynamics relies upon mathematical models informed by scientific knowledge and relevant data.
The models aim to provide a statistical description of the trajectory of an epidemic and its uncertainty, together with a representation of the underlying biological mechanisms.
Evaluation of success at these goals is necessary in order for a model to provide a reliable tool for guiding evidence-based public policy interventions.
\old{In this article, we discuss common modeling decisions that may not provide an adequate statistical explanation of the data, and we show that attention to statistical fit can lead to more accurate policy evaluations}.
\new{In this article, we conduct a re-analysis of the 2010 cholera outbreak in Haiti. We use this case study to investigate current procedures for fitting mathematical models to time series data, while identifying limitations of these methodologies and proposing remedies.}
\new{We present} \old{Our analysis presents} methodology for diagnosing how well a model describes observed data\new{.}\old{, and provides careful justification of valid conclusions from the fitted model.}
Objective measures are used to \old{benchmark} \new{evaluate} model fit; when \old{these}\new{objective measures of the goodness-of-fit of a model} are combined with reproducibility, a framework emerges for continual improvement \new{to model based inference} when revisiting the data\old{ and models}.
\new{This framework promotes scientific discovery and supports model-driven policy recommendations by simplifying the process of understanding and improving previous results.}
Our data analysis workflow is supported by recent advances in algorithms, software and hardware, which facilitate statistical fitting of nonlinear stochastic dynamic models to observed incidence data.
However, inference for high-dimensional systems remains a methodological challenge.
One of the models under consideration involves spatially coupled stochastic meta-populations, and we demonstrate how a recently developed algorithm permits likelihood-based inference and model diagnostics in this setting.
We contend that raising the currently accepted standards of infectious disease modeling will result in a greater ability of scientists and policy makers to understand and respond to future infectious disease outbreaks.

@@

This case study investigates the use of dynamic models and spatiotemporal data to inform \old{a policy decision} \new{public health policy} in the context of the cholera outbreak in Haiti, which started in 2010.
\old{We build on a multi-group modeling exercise by Lee et al. [1] in which four expert modeling teams developed models to the same dataset with the goal of comparing conclusions on the feasibility of eliminating cholera by a vaccination campaign}.
\old{The data that were used for this study, and that is the subject of our reanalysis, is displayed in Fig.~[1].}
\new{Various dynamic models were developed to study this outbreak: searching PubMed with keywords ``Haiti, cholera, model'' we obtained 22 studies that utilized deterministic mechanistic dynamic models \cite{lee20,tuite11,andrews11,botelho21,fitzgibbon20,eisenberg13,rinaldo12,chao11,date11,lin19,abrams13,collins21,akman15,trevisin22,mavian20,collins14,kelly16,capone15,leung22,mari15,gatto12,kuhn14}
and 11 studies that used stochastic models \cite{kirpich16,lee20,pasetto18,mukandavire13,kirpich17,lewnard16,kunkel17,mukandavire15,sallah17,azman12,azman15}.
Incidence data on the outbreak are available at various spatial scales, motivating 17 studies in our literature review to consider spatially explicit dynamic models \cite{lee20,tuite11,pasetto18,fitzgibbon20,eisenberg13,rinaldo12,chao11,abrams13,trevisin22,sallah17,collins14,kelly16,azman12,leung22,kuhn14,mari15,gatto12}.
Here we focus on a multi-group modeling exercise by Lee et al.~\cite{lee20} in which four expert modeling teams developed models to the same dataset with the goal of comparing conclusions on the feasibility of eliminating cholera by a vaccination campaign.}
Model~1 is stochastic and describes cholera at the national level;
Model~2 is deterministic with spatial structure, and includes transmission via contaminated water;
Model~3 is stochastic with spatial structure, and accounts for measured rainfall.
Model~4 has an agent-based construction, featuring considerable mechanistic detail but limited ability to calibrate these details to data.
\new{These modeling strategies were selected by Lee et a.~\cite{lee20} to represent the range of approaches used in the research community}
We focus on Models~1--3, as the strengths and weaknesses of the agent-based modeling approach \cite{tracy18} are outside the scope of this article.
\new{In addition, agent-based models were less widely used, the agent based model in \cite{lee20} being the only model of this class that was found in our literature review.}
\new{The data that were used in \cite{lee20}, and that we reanalyze, are displayed in Fig.~\ref{fig:CholeraData}.}

@@

The four independent teams were given the task of estimating the
potential effect of prospective oral cholera vaccine (OCV) programs.
While OCV is accepted as a safe and effective tool for controlling the spread of cholera, the global stockpile of OCV doses remains limited \cite{pezzoli20}.
Advances in OCV technology and vaccine availability, however, raised the possibility of planning a national vaccination program.
\new{The possibility of controlling the Haiti cholera outbreak via OCV was considered by various research groups \cite{lee20,ivers17,andrews11,walton11,matias17,chao11,fung13,date11,leung22,azman15}}.
In the \new{Lee et al.~\cite{lee20}} study, certain data were shared between the groups, including demography and vaccination history; vaccine efficacy was also fixed at a shared value between groups.
Beyond this, the groups made autonomous decisions on what to include and exclude from their models.
\old{The groups largely adhered to existing guidelines for creating models to inform policy [35, 36] and, d}\new{D}espite their autonomy, \new{the four independent teams} obtained a consensus that an extensive nationwide vaccination campaign would be necessary to eliminate cholera from Haiti.
\new{They also provided a range for the forecasted number of cumulative cases from February 2019 to Februaruy 2024 based on model simulations, with median estimates from each model measured in hundreds of thousands.}
\old{Their forecasts, however, were} \new{These forecasts are} inconsistent with the prolonged period with no confirmed cholera cases between February, 2019 and September, 2022 \cite{trevisin22}.
Though cholera has recently reemerged in Haiti \cite{rubin22,PAHO23}, the inability to accurately forecast cholera incidence from 2019-2022\old{---while diligently adhering to existing scientific standards for infectious disease modeling---demonstrates that existing standards of disease modeling and forecasting may be insufficient.} \new{prompts us to consider retrospectively what may have been done differently in order to obtain more reliable conclusions.}

@@

\new{Confidence intervals for model parameters are given in supplement~S8.} %TODO: nameref this

@@

\subsubsection*{Model~1}
\label{sec:model1}
The latent state vector $\bm{X}(t) = \big(S_{\vaccCounter}(t),  E_{\vaccCounter}(t), I_{\vaccCounter}(t), A_{\vaccCounter}(t), R_{\vaccCounter}(t), \vaccCounter \in 0:\vaccClass\big)$ describes susceptible, latent (exposed), infected (and symptomatic), asymptomatic, and recovered individuals in vaccine cohort $\vaccCounter$ at time $t$.
Here, $\vaccCounter=0$ corresponds to unvaccinated individuals, and $\vaccCounter \in \seq{1}{\vaccClass}$ describes hypothetical vaccination programs. \new{Each program $\vaccCounter$ indexes differences in both the number of doses administered (one versus two doses per individual) and the round of vaccine administration, seperating individuals into compartments with distinct dynamics based on vaccination status.}
The force of infection is
\begin{equation}
\label{model1:lambda}
\lambda(t) = \Big(\sum_{\vaccCounter=0}^{\vaccClass}  I_{\vaccCounter}(t) + \asymptomRelativeInfect \sum_{\vaccCounter=0}^{\vaccClass} A_{\vaccCounter}(t) \Big)^\nu \frac{d\Gamma(t)}{dt} \transmission(t)/N,
\end{equation}
where $\transmission(t)$ is a periodic cubic spline representation of seasonality, given in terms of a B-spline basis $\{ s_j(t), j\in \seq{1}{6}\}$ and parameters $\transmission_{1:6}$ as
\begin{equation}
\label{model1:beta}
\log\transmission(t) = \sum_{j=1}^6 \transmission_j s_j(t).
\end{equation}
The process noise $d\Gamma(t)/dt$ is multiplicative Gamma-distributed white noise, with infinitesimal variance parameter $\sigmaProc^2$.
Lee et al.~\cite{lee20} included process noise in Model~3 but not in Model~1, i.e., they fixed $\sigmaProc^2 = 0$.
Gamma white noise in the transmission rate gives rise to an over-dispersed latent Markov process \cite{breto11} which has been found to improve the statistical fit of disease transmission models \cite{stocks20,he10}.

@@

Individuals move from department $u$ to $v$ at rate $T_{uv}$, and aquatic cholera moves at rate $T^W_{uv}$.
The nonzero transition rates are
\begin{eqnarray}
\label{model2:mu_SE}
\mu_{S_{u\vaccCounter}E_{u\vaccCounter}} &=& \new{(1 - \vaccineEfficacy_\vaccCounter) \, \lambda_u(t)},
\\
\label{model2:mu_EI}
\mu_{E_{u\vaccCounter}I_{u\vaccCounter}} &=& \symptomFrac\muEI, \quad \mu_{E_{u\vaccCounter}A_{u\vaccCounter}} = (1-\symptomFrac)\muEI,
\\
\label{model2:mu_IR}
\mu_{I_{u\vaccCounter}R_{u\vaccCounter}} &=& \mu_{A_{u\vaccCounter}R^A_{u\vaccCounter}} = \muIR,
\\
\label{model2:RS}
\mu_{R_{u\vaccCounter}S_{u\vaccCounter}} &=& \mu_{R^A_{u\vaccCounter}S_{u\vaccCounter}} = \muRS,
\\
\label{model2:transport}
\mu_{S_{u\vaccCounter}S_{{\varv}\vaccCounter}} &=& \mu_{E_{u\vaccCounter}E_{{\varv}\vaccCounter}} = \mu_{I_{u\vaccCounter} I_{{\varv}\vaccCounter}} = \mu_{A_{u\vaccCounter}A_{{\varv}\vaccCounter}} = \mu_{R_{u\vaccCounter}R_{{\varv}\vaccCounter}} = \mu_{R^A_{u\vaccCounter} R^A_{{\varv}\vaccCounter}} = T_{u\varv},
\\
\label{model2:omega1}
\mu_{S_{u1}S_{u0}} &=& \mu_{S_{u3}S_{u0}} = \omega_1,
\\
\label{model2:omega2}
\mu_{S_{u2}S_{u0}} &=& \mu_{S_{u4}S_{u0}} = \omega_2,
\\
\label{model2:to_W}
\mu_{\demography W_u} &=& \Wshed \left\{ \sum_{\vaccCounter=0}^4 I_{u\vaccCounter}(t) + \asymptomRelativeShed \sum_{\vaccCounter=0}^4 A_{u\vaccCounter}(t) \right\},
\\
\label{model2:from_W}
\mu_{W_u\demography} &=& \Wremoval,
\\
\label{model2:water_transport}
\mu_{W_uW_{\varv}} &=& w_r T^W_{u\varv}.
\end{eqnarray}
In Eq.~\myeqref{model2:transport} the spatial coupling is specified by a gravity model,
\begin{equation}
\label{model2:gravity}
T_{u\varv} = v_{\mathrm{rate}} \times \frac{\mathrm{Pop}_u \mathrm{Pop}_{\varv}}{D_{u\varv}^2},
\end{equation}
where $\mathrm{Pop}_u$ is the mean population for department $u$,
$D_{u\varv}$ is a distance measure estimating average road distance between randomly chosen members of each population, and $v_{\mathrm{rate}}= 10^{-12} \,\mbox{km$^2$yr$^{-1}$}$ \old{was treated as a fixed constant.}\new{was fixed at the value used in \cite{lee20}.}
In Eq.~\myeqref{model2:water_transport}, $T^W_{u\varv}$ is a measure of river flow between departments.
The unit of $W_u(t)$ is cells per ml, with dose response modeled via a saturation constant of $\Wsat$ in Eq.~\myeqref{model2:lambda}.
In Eq.~\myeqref{model2:mu_SE}, $\vaccineEfficacy_\vaccCounter$ denotes the vaccine efficacy for each vaccination campaign $\vaccCounter \in \vaccClass$, with \old{$\vaccineEfficacy_0 = 1$, $\vaccineEfficacy_1 = 1 - 0.429 \childReduce$, $\vaccineEfficacy_2 = 1 - 0.519 \childReduce$, $\vaccineEfficacy_3 = 1 - 0.429$, and $\vaccineEfficacy_4 = 1 - 0.519$} \new{$\vaccineEfficacy_0 = 0$, $\vaccineEfficacy_1 = 0.429 \childReduce$, $\vaccineEfficacy_2 = 0.519 \childReduce$, $\vaccineEfficacy_3 = 0.429$, and $\vaccineEfficacy_4 = 0.519$}
Here, $\childReduce = 0.4688$ represents the reduced efficacy of the vaccination for children under the age of five years, and the values $0.429$ and $0.519$ are the median effectiveness of one and two doses over their effective period respectively, according to Table S4 in the supplement material of Lee et al.~\cite{lee20}.
Because vaccine efficacy remains constant, individuals in this model transition from a vaccinated compartment to the susceptible compartment at the end of the vaccine coverage period.

@@

\begin{table}[!h]
\centering
\caption{\label{tab:likes}{\bf AIC values for Models~1--3 and their benchmarks.}}
\begin{tabular}{|c|c|c|c|}
\hline
 & \thead{Model~1} & \thead{Model~2} & \thead{Model~3}
\\
\hline
\hline
\multirow{2}{*}{Log-likelihood} &
  $\Sexpr{myround(mod1_ll, 1)}$ &
  $\Sexpr{myround(mod2_ll, 1)}$ &
  $\Sexpr{myround(mod3_ll, 1)}$ \\
    & ($\Sexpr{myround(mod1_lee_ll, 1)}$)\footnotemark[1] &
  ($\Sexpr{myround(unname(lee20_replication['leeFit_ll']), 1)}$) &
  ($\Sexpr{myround(lee3_results[["all_ll_eval"]], 1)}$)\footnotemark[2]
\\
\hline
Number of &
  $\Sexpr{as.character(mod1_n_params)}$ &
  $\Sexpr{as.character(mod2_n_params)}$ &
  $\Sexpr{as.character(mod3_n_params)}$ \\
 Fit Parameters &
 (\Sexpr{as.character(mod1_lee_n_params)}) &
 ($\Sexpr{as.character(unname(lee20_replication['leeFit_n_params']))}$) &
 ($\Sexpr{as.character(lee3_results$n_params, 1)}$)
\\
\hline
\multirow{2}{*}{AIC} &
  $\Sexpr{myround(mod1_aic, 1)}$ &
  $\Sexpr{myround(mod2_aic, 1)}$ &
  $\Sexpr{myround(mod3_aic, 1)}$ \\
  & ($\Sexpr{myround(mod1_lee_aic, 1)}$)\footnotemark[1] &
  ($\Sexpr{myround(mod2_lee_aic, 1)}$) &
  ($\Sexpr{myround(mod3_lee_coupled_aic, 1)}$)\footnotemark[2]
\\
\hline
\thead{Benchmark \\ AIC} &
  $\Sexpr{myround(-2 * benchmarks$m1 + 2 * 4, 1)}$ &
  $\Sexpr{myround(-2 * benchmarks$m2 + 2 * 40, 1)}$ &
  $\Sexpr{myround(-2 * benchmarks$m3 + 2 * 40, 1)}$
\\
\hline
\end{tabular}
\begin{flushleft}
Values in parentheses are corresponding values \new{obtained} using \new{the models of \cite{lee20}} \old{Lee et al.~[4] parameter estimates}.  \textsuperscript{1}The reported likelihood is an upper bound of the likelihood of the \old{Lee et al.~[4]} model \new{in \cite{lee20}}. \old{Lee et al.~[4] fit} \textsuperscript{2}\new{In \cite{lee20},} Model~3 \new{was fit} to a subset of the data (March 2014 onward, excluding data from Ouest in 2015-2016).
On this subset, their model has a likelihood of $\Sexpr{myround(lee3_results$sub_ll_eval, 1)}$.
On this same subset, our model has a likelihood of $\Sexpr{myround(mod3_eval_res$subset_ll['est'], 1)}$.
Details of estimating the likelihood of the \old{Lee et al.~[4]} models \new{used in \cite{lee20} are} \old{is} provided in \nameref{S_lee20}
\old{See Sec.~S6 of the supplement material for more details on estimating the likelihood of the Lee et al.~[4] models.}
\end{flushleft}
\end{table}

@@

@@

The ongoing global COVID-19 pandemic has \old{provided a clear example on} \new{demonstrated} how government policy may be affected by the \old{conclusions of scientific models} \new{inferences drawn from mathematical modeling} \cite{saltelli20}.
\old{This article demonstrates that fitting appropriate scientific models to guide policy is a challenging statistical task.}
\new{However, the development of credible models---which are supported by data and can lend quantitative insights into a dynamic system---remains a challenging task.
In this article, we demonstrate opportunities available for raising the current standards of statistical inference for mathematical models of biological systems.}
